RED WINE QUALITY DATA EXPLORATION by SIDDHARTH SHANKAR
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Importing required libraries
library(ggplot2)
library(gridExtra)
library(corrplot)
library(GGally)
library(Hmisc)
library(memisc)

# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
red_wine = read.csv('wineQualityReds.csv')
```

###Abstract
I will analyze the Red Wine Dataset. Key goals of the study are to understand
which chemical properties influence the quality of red wines and its correlation
among them.

###Introduction
**About the data**: The red wine data set contains 1,599 red wines with 11 
variables on the chemical properties of the wine. At least 3 wine experts rated 
the quality of each wine, providing a rating between 0 (very bad) and 10 
(very excellent). \
Number of Attributes: 11 + output attribute \
Attribute information: \
Input variables (based on physicochemical tests): 

1. fixed acidity (tartaric acid - g / dm^3)
2. volatile acidity (acetic acid - g / dm^3)
3. citric acid (g / dm^3)
4. residual sugar (g / dm^3)
5. chlorides (sodium chloride - g / dm^3
6. free sulfur dioxide (mg / dm^3)
7. total sulfur dioxide (mg / dm^3)
8. density (g / cm^3)
9. pH
10. sulphates (potassium sulphate - g / dm3)
11. alcohol (% by volume) \
Output variable (based on sensory data): \
12. quality (score between 0 and 10)

####Description of attributes:

1. **fixed acidity**: most acids involved with wine or fixed or nonvolatile 
(do not evaporate readily)
2. **volatile acidity**: the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste
3. **citric acid**: found in small quantities, citric acid can add 'freshness' 
and flavor to wines
4. **residual sugar**: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater than 
45 grams/liter are considered sweet
5. **chlorides**: the amount of salt in the wine
6. **free sulfur dioxide**: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine
7. **total sulfur dioxide**: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
8. **density**: the density of water is close to that of water depending on the 
percent alcohol and sugar content
9. **pH**: describes how acidic or basic a wine is on a scale from 0 
(very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale
10. **sulphates**: a wine additive which can contribute to sulfur dioxide gas 
(S02) levels, wich acts as an antimicrobial and antioxidant
11. **alcohol**: the percent alcohol content of the wine \
Output variable (based on sensory data): \
12. **quality (score between 0 and 10)**

# Univariate Plots Section

#### Brief Data & Summary of the dataset

```{r echo=FALSE, Univariate_Plots}
# Reading the first few records
head(red_wine)
# Print summary statistics about the dataset
summary(red_wine)

```

#### Wine Quality Distribution

```{r echo=FALSE, Wine_Quality_Distribution}
# Plotting bar chart for discrete variable, quality
ggplot(data = red_wine, aes(quality)) +
  geom_bar(colour="black", fill="#722F37",stat = "count") +
                xlab("Wine Quality") + ylab("Counts") +
                geom_vline(aes(xintercept = mean(quality)),col='blue',size=1) +
                geom_vline(aes(xintercept = median(quality)),col='red',size=1)

```

Wine quality is a categorical discrete variable and it ranges from 3 to 8 in 
the given dataset. There are exceptionally no good or bad wines. Treating the 
data as continuous will give the mean as <span style="color:blue"> 5.636 </span> and median as <span style="color:red"> 6 </span>.

#### Chemical Properties Distribution

```{r echo=FALSE, Wine_Chemical_Properties}
# Function to display histogram plot
geom_histogram_plot <- function(properties, xlabel, val) {
  return (ggplot(data = red_wine, aes(x = properties)) +
  geom_histogram(colour="black", fill="#722F37", binwidth = val ) +
                xlab(xlabel) + ylab("Counts")) }
# Store the plot outputs in variables
p1 <- geom_histogram_plot(red_wine$fixed.acidity,"Fixed Acidity",0.5)
p2 <- geom_histogram_plot(red_wine$volatile.acidity,"Volatile Acidity",0.05)
p3 <- geom_histogram_plot(red_wine$citric.acid,"Citric Acid",0.05)
p4 <- geom_histogram_plot(red_wine$residual.sugar,"Residual Sugar",0.5)
p5 <- geom_histogram_plot(red_wine$chlorides,"Chlorides",0.025)
p6 <- geom_histogram_plot(red_wine$free.sulfur.dioxide,"Free Sulfur Dioxide",4)
p7 <- geom_histogram_plot(red_wine$total.sulfur.dioxide,"Total Sulfur Dioxide",
                          10)
p8 <- geom_histogram_plot(red_wine$density,"Density",0.001)
p9 <- geom_histogram_plot(red_wine$pH,"pH",0.05)
p10 <- geom_histogram_plot(red_wine$sulphates,"Sulphates",0.1)
p11 <- geom_histogram_plot(red_wine$alcohol,"Alcohol",0.25)

print(p1)
```

Fixed Acidity appears to be largely positively skewed with a mean of 8.32 and 
median 7.90.

```{r echo=FALSE}
print(p9)
```

The  mean and median of pH are approximately equal with the values of 3.311 and 
3.310 respectively which denotes that pH is normally distributed. Also, a little 
research online showed that red wines has a pH value range from 3.3 to 3.6.

```{r echo=FALSE, Wine_Chemical_Properties_1}
# Using the stored plot variables to arrange the plots in a grid
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11, ncol=3)
```

# Univariate Analysis
We plotted the histograms of 11 different chemical properties of red wine to 
get an idea of the dispersion of each properties. Based on the histograms 
plotted above, the following observations can be made on the distribution of 
chemical properties:

1. **Normally Distributed**: Volatile Acidity, Density, pH
2. **Positively Skewed**: Fixed Acidity, Citric Acid, Free Sulfur Dioxide, 
Total Sulfur Dioxide, Sulphates, Alcohol
3. **Long Tail**: Residual Sugar, Chlorides

Large outliers can be seen in positively skewed and long tailed variables. 
We will transform some of them to normal distribution by taking log10 which
will produce a relatively normal distribution.

```{r echo=FALSE, Transform_Long_Tailed}
# Log10 transformations to get a relatively normal distribution
p1 <- ggplot(data = red_wine, aes(x = log10(residual.sugar))) +
  geom_histogram(colour="black", fill="#722F37", binwidth = 0.05 ) +
                xlab("log10 of Residual Sugar") + ylab("Counts")

p2 <- ggplot(data = red_wine, aes(x = log10(chlorides))) +
  geom_histogram(colour="black", fill="#722F37", binwidth = 0.05 ) +
                xlab("log10 of Chlorides") + ylab("Counts")

p3 <- ggplot(data = red_wine, aes(x = log10(total.sulfur.dioxide))) +
  geom_histogram(colour="black", fill="#722F37", binwidth = 0.05 ) +
                xlab("log10 of Total Sulfur Dioxide") + ylab("Counts")

p4 <- ggplot(data = red_wine, aes(x = log10(sulphates))) +
  geom_histogram(colour="black", fill="#722F37", binwidth = 0.02 ) +
                xlab("log10 of Sulphates") + ylab("Counts")
# Arrange the transformations in a grid
grid.arrange(p1,p2,p3,p4, ncol = 2)
```

### What is the structure of your dataset?
There are 1,599 observations with 11 attributes (11 variables on the chemical 
properties of the wine) + 1 output attribute (quality of red wine).

### What is/are the main feature(s) of interest in your dataset?
The quality rating is the main feature in the dataset which defines the good 
and bad taste of the red wine.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Based on the above distributions, I think that fixed acidity, citric acid, 
residual sugar, pH, chlorides will be the features of interest.

### Did you create any new variables from existing variables in the dataset?
No.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Some of the distributions were positively skewed and long tailed which I have 
transformed to produce a relatively normal distribution.

# Bivariate Plots Section

Let's begin with examining the correlation between two variables using 
correlation plot.

```{r echo=FALSE, Bivariate_Plots_1, fig.width=9, fig.height=8}
# Since, quality is a categorical data, converting it as a numeric
# To plot Correlation Matrix
red_wine$numquality <- as.numeric(red_wine$quality)

corr_matrix <- cor(red_wine[c(2:13, 14)])

corrplot(corr_matrix, addCoef.col = "black", 
         order = "hclust", insig = "pch", addrect = 3)
```

#### Dataset of correlation coefficient for Bivariates
```{r echo=FALSE, Bivariate_Plots_2}
# Create a correlation matrix with decimal places upto 3 digits 
z <- round(cor(red_wine),3)
# Convert the matrix in to a dataframe 
zdf <- as.data.frame(as.table(z))
# Taking the subset of the dataframe having frequencies greater than 0.5
z1 <- subset(zdf, abs(Freq) > 0.5)
print(z1[order(abs(z1$Freq)),])
```

The top 4 chemical properties that are correlated are:

1. fixed acidity & pH with the correlation coefficient of -0.683 stating that
pH tends to decrese with increase in fixed.acidity

```{r echo=FALSE, Bivariate_Plots_2_1}
# Create scatterplot for confirming relation with fixed acidity and pH
ggplot(red_wine, aes(x = fixed.acidity, y = pH)) +
  geom_point(color = "#722F37", alpha = 1/4) + 
  stat_smooth(method = 'lm')

cor(red_wine$fixed.acidity, red_wine$pH)
```


2.citric acid & volatile acidity with the correlation coefficient of -0.552
stating that citric acid tends to decrease with increase in volatile acidity

```{r echo=FALSE, Bivariate_Plots_2_2}
# Create scatterplot for confirming relation with citric acid and volatile acidity
ggplot(red_wine, aes(x = citric.acid, y = volatile.acidity)) +
  geom_point(color = "#722F37", alpha = 1/4) + 
  stat_smooth(method = 'lm')

cor(red_wine$citric.acid, red_wine$volatile.acidity)
```

3.citric acid & pH with the correlation coefficient of -0.542 (slightly weaker) 
stating that pH tends to decrease with increase in citric acid

```{r echo=FALSE, Bivariate_Plots_2_3}
# Create scatterplot for confirming relation with citric acid and pH
ggplot(red_wine, aes(x = citric.acid, y = pH)) +
  geom_point(color = "#722F37", alpha = 1/4) + 
  stat_smooth(method = 'lm')

cor(red_wine$citric.acid, red_wine$pH)
```

4. citric acid & fixed acidity with the correlation coefficient of 0.672 stating
that fixed acidity increases with increase in citric acid

```{r echo=FALSE, Bivariate_Plots_2_4}
# Create scatterplot for confirming relation with citric acid & fixed acidity
ggplot(red_wine, aes(x = citric.acid, y = fixed.acidity)) +
  geom_point(color = "#722F37", alpha = 1/4) + 
  stat_smooth(method = 'lm')

cor(red_wine$citric.acid, red_wine$fixed.acidity)
```

It can be observed that citric acid is a subset of fixed acidity. 

Let us now abserve the boxplots of the selected variables and its median will 
give a better measure of variance in the dataset.

```{r echo=FALSE, Bivariate_Plots_3_1}
# Create a new column with wine_grade
red_wine$wine_grade <- cut(red_wine$quality, c(2.5,3.5,4.5,5.5,6.5,7.5,8.5),
                          labels=c('3','4','5','6','7','8'))
# Create boxplot for alcohol vs wine grade
ggplot(red_wine, aes(x = wine_grade, y = alcohol)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Alcohol") +
       ggtitle("Alcohol vs Wine Grade")
```

Higher quality wine tend to have higher alcholol content as compared to low 
quality wines.

```{r echo=FALSE, Bivariate_Plots_3_2}
# Create boxplot for volatile acidity vs wine grade
ggplot(red_wine, aes(x = wine_grade, y = volatile.acidity)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Volatile Acidity") +
       ggtitle("Volatile Acidity vs Wine Grade")
```

Volatile acidity decreases as the wine grades increases. Volatile acidity is 
responsible for the smell in wine and too much of it will reduce the wine quality.

```{r echo=FALSE, Bivariate_Plots_3_3}
# Create boxplot for citric acid vs wine grade
ggplot(red_wine, aes(x = wine_grade, y = citric.acid)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Citric Acid") +
       ggtitle("Citric Acid vs Wine Grade")
```

Citric acid greatly affects the quality of wine. In low grade red wines, its 
median is almost pointing to 0 while a well balanced citric acid increases the
quality of wine.

```{r echo=FALSE, Bivariate_Plots_3_4}
# Create boxplot for sulphates vs wine grade
ggplot(red_wine, aes(x = wine_grade, y = sulphates)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Sulphates") +
       ggtitle("Sulphates vs Wine Grade")
```

Though sulphates are used to maintain the freshness of wines, higher the presence
of sulphates in wines, increases the wine graded.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
Volatile acidity is responsible for the aroma of wine and is not intentionally
included in the wine. It can be observed from the boxplot of volatile acidity 
and wine grade that higher the volatile acidity, lower is the quality of wine 
and vice-versa. Also, Higher quality of wine tends to have high level of 
alcohol. The median for sulphates increases for each wine grade (quality).

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
When citric acid increases, fixed acidity also increases denoting a positive 
correlation. Citric acid and volatile acidity are negatively correlated. Citric 
acid and pH were also negatively correlated – a lower pH indicates a higher 
acidity. 


### What was the strongest relationship you found?
pH & Fixed Acidity with the correlation coefficient of -0.683.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots_1_1}
# scatterplot (jittered) - quality and alcohol, by volatile acidity
ggplot(red_wine, aes(x=wine_grade,y=alcohol, color = volatile.acidity)) +
        geom_jitter(height = 0, width = 0.35, alpha=0.8) +
        scale_colour_gradient2(low = "#40e0d0", high = "#FF0080",
                               midpoint = median(red_wine$volatile.acidity))  +
        ggtitle("Quality and Alcohol, by Volatile Acidity") +
  theme(panel.background = element_rect(fill = "darkgrey"))
```

It can be observed that higher quality wine has lower volatile acidity.

```{r echo=FALSE, Multivariate_Plots_1_2}
# scatterplot - alcohol and volatile acidity, by quality level
ggplot(red_wine, aes(x = alcohol, y = volatile.acidity, 
                     color = as.factor(quality))) +
        geom_point(alpha=0.8) + 
  scale_color_brewer(palette = "RdYlBu", name="Quality", direction=-1) +
  theme(panel.background = element_rect(fill = "darkgrey")) +
  ggtitle("Alcohol and Volatile Acidity, by Quality level")
```


```{r echo=FALSE, Multivariate_Plots_1_3}
# Scatterplot (jittered) alcohol and volatile acidity by log of sulphates
ggplot(red_wine, aes(x = alcohol, y = volatile.acidity, 
        color = log10(red_wine$sulphates))) +
        geom_jitter(alpha=1/2,size=3) +
        scale_colour_gradient2(low = "#40e0d0", mid = "#FF8C00", 
                               high = "#FF0080", midpoint = 0) +
        facet_grid(wine_grade ~ .)
```

It can be observed that higher quality wine have higher alcohol, lower 
volatile acidity and higher sulphates.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
The multivariate analysis only stregthen the relationship we observed in the 
bivariate analysis. It depicts that higher quality wine have higher alcohol, 
lower volatile acidity and higher sulphates.

### Were there any interesting or surprising interactions between features?
No

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Plotting bar chart for discrete variable, quality
ggplot(data = red_wine, aes(quality)) +
  geom_bar(colour="black", fill="#722F37",stat = "count") +
                xlab("Wine Quality") + ylab("Counts") +
                ggtitle("Distribution of Wine Quality") +
                geom_vline(aes(xintercept = mean(quality)),col='blue',size=1) +
                geom_vline(aes(xintercept = median(quality)),col='red',size=1)
                
```

### Description One
Plot one shows the distribution of wine quality based on the physicochemical 
tests. It can be observed that the given dataset of red wine contains a large 
number of wines that are average in quality. The mean and median of the quality 
of red wines are 5.636 and 6 respectively.

### Plot Two
```{r echo=FALSE, Plot_Two}
# Create boxplots for variables of interests
p1 <- ggplot(red_wine, aes(x = wine_grade, y = alcohol)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Alcohol (% by volume)") +
       ggtitle("Alcohol vs Wine Grade")

p2 <- ggplot(red_wine, aes(x = wine_grade, y = volatile.acidity)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Volatile Acidity (g / dm^3)") +
       ggtitle("Volatile Acidity vs Wine Grade")

p3 <- ggplot(red_wine, aes(x = wine_grade, y = citric.acid)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Citric Acid (g / dm^3)") +
       ggtitle("Citric Acid vs Wine Grade")

p4 <- ggplot(red_wine, aes(x = wine_grade, y = sulphates)) + 
       geom_boxplot(color = "darkblue") + guides(fill = FALSE) +
       xlab("Wine Grade") + ylab("Sulphates (g / dm3)") +
       ggtitle("Sulphates vs Wine Grade")
# Arranging the boxplots in a gird
grid.arrange(p1,p2,p3,p4, ncol = 2)
```

### Description Two
Based on the correlation, the following 4 chemcial properties have the highest 
correlation coefficient: Alcohol, Volatile Acidity, Citric Acid &Sulphates. 
Higher the wine grade, higher is the level of alcohol and citric acid. If we
group wine grades as bad (3,4), average (5,6) and good (7,8), we can observe 
that average wines have higher content of sulphates and alcohol in it. Also, the
level of sulphates increases slightly in good grade wines which acts as an 
important role in maintaining the freshness of the wine. 

### Plot Three
```{r echo=FALSE, Plot_Three}
# scatterplot - alcohol and volatile acidity, by quality level
ggplot(red_wine, aes(x = alcohol, y = volatile.acidity, 
                     color = as.factor(quality))) +
        geom_point(alpha=0.8) + 
  scale_color_brewer(palette = "RdYlBu", name="Quality", direction=-1) +
  theme(panel.background = element_rect(fill = "darkgrey")) +
  ggtitle("Alcohol and Volatile Acidity, by Quality") + 
  xlab("Alcohol (% by Volume)") +
  ylab(expression(Volatile~Acidity~(g/dm^{3})))
```

### Description Three
It can be observed that higher quality wine has more alcohol content and less
volatile.acidity which means that the quality of wine increases with the 
increase in alcohol and decrease in volatile acidity.

# Reflection
The key goals of this study were to understand which chemical properties 
influence the quality of red wines and its correlation among them. The red wine 
data set contains 1,599 red wines with 11 variables on the chemical properties 
of the wine. Initially, when I plotted the histograms of all the 11 variables,
based on the nature of the plots, I assumed that some of these variables are 
related to each other like being directly or inversely proportional or subset 
which turns out to be true in the correlation analysis. The correlation showed 
that pH tends to decrease with increase in fixed acidity and citric acid and 
fixed acidity goes hand in hand i.e. they are positively correlated with a value
of 0.6717. After doing some web research I learned about a few things about the 
presence of different chemical properties in wine. 

Volatile acidity has a negative correlation. It refers to the acidic elements 
of the wine that are gaseous rather than liquid. It is the acetic acid compound 
which is majorly responsible for the aroma. Though it is not intentionally 
included in the wine,but is an important characterstic in many wines that adds 
complexity and interest; often in positive manner. 

Presence of alcohol plays an important role in determining the quality of wines.
Wines having higher level of alcohol provides rich, ripe fruits flavors. Those 
flavors come from really ripe grapes, and really ripe grapes come from warmer 
growing conditions.Those grapes contain more sugar, and more sugar produces 
more alcohol during fermentation.

The presence of sulphates in wine determine its freshness and based on the 
correlation the level of sulphates increases with increase in wine quality.

Further improvements can be done, if data for exceptionally good and bad wines 
are present. However, examining the quality of wine is complex and therefore, 
apart from chemical properties if more factors such as storage duration, 
quality and types of grapes, etc. are provided the quality of analysis can be 
improved.
 

### Problems & Solutions
When I plotted the correlation matrix, all the data were overlapped and it 
looked messy. A google search showed how to show the data as an ordered list 
and then I created a correlation matrix, transformed it into a dataframe and
ordered the data above a certain value to show only relevent values.


# References
[1] https://stackoverflow.com/questions/7074246/show-correlations-as-an-ordered-list-not-as-a-large-matrix/7074856

[2] https://www.decanter.com/learn/volatile-acidity-va-45532/

[3] https://www.thekitchn.com/the-truth-about-sulfites-in-wine-myths-of-red-wine-headaches-100878

[4] https://www.tennessean.com/story/life/food/2015/04/17/alcohol-content-affect-wine/25779589/